---
title: "Bills of Material"
subtitle: ""
summary: ""
author: "Eric Neer"
output: 
  html_document:
    toc: true
date: 2020-06-01
categories: []
tags: []
projects: []
draft: true
references:
- id: lunn1992
  title: "MRP: Integrating Material Requirements Planning and Modern Business"
  author:
    - family: Lunn
      given: Terry
    - family: Neff
      given: Susan
  issued:
    year: 1992
  type: book
---

```{r, include=FALSE}
library(DiagrammeR)
library(RSQLite)
```

### Description
In manufacturing, a bill of materials defines all of the items required to produce a good.
Often times this consists of raw materials (e.g. steel, aluminum) that are transformed or other components that are assembled to make the finished product.
It may also contain ancillary items used during manufacturing, such as tooling or consumable chemicals.

<div align="center">
```{r echo=FALSE}
grViz("
  digraph {
    node [shape='box']
    edge [arrowhead='none']
    LP [label='Lower Pen']
    UB [label='Upper Barrel']
    T1 [label='Tubing']
    T2 [label='Tubing']
    
    Pen -> {LP Ink UB}
    LP -> {Inner Point Outer}
    Inner -> T1
    Outer -> T2
    UB -> {Sleeve Clip}
    Sleeve -> {Top Tube}
    Tube -> Tubing
    Clip -> Steel
    
    node [shape='plaintext']
    L0 [label='Level 0']
    L1 [label='Level 1']
    L2 [label='Level 2']
    L3 [label='Level 3']
    L4 [label='Level 4']
    
    subgraph labels {
      rankdir='TB'
      L0 -> L1 -> L2 -> L3 -> L4 [style='invis']
    }
  }
")
```
<figcaption>Hypothetical multi-level bill of materials for a pen. [@lunn1992]</figcaption>
</div>


### Data Structure
+ Adjacency lists
+ Graphs

### References